---
title: "Polymerase fidelity estimates"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required libraries, load and merge MAGERI results.

```{r}
library(plyr); library(ggplot2); library(reshape2); library(gplots); library(knitr); library(RColorBrewer)

load_variant_table <- function(path, project, sample) {
  fname <- paste(path, paste(project, sample, "variant.caller.txt", sep ="."), sep = "/")
  df.1 <- read.table(fname, header=T, sep="\t")
  df.1$project <- project
  df.1$sample <- sample
  df.1
}

# mutation signatures
sign.rep <- data.frame(mutation.signature = c("A>C","A>G","A>T","C>A","C>G","C>T","G>A","G>C","G>T","T>A","T>C","T>G"),
                       mutation.signature.rep = c("A>C,T>G","A>G,T>C","A>T,T>A","C>A,G>T","C>G,G>C","C>T,G>A","C>T,G>A","C>G,G>C","C>A,G>T","A>T,T>A","A>G,T>C","A>C,T>G"))

path <- "data/"
df.meta <- read.table(paste(path, "metadata.txt", sep="/"), header=T, sep = "\t")

# load and concatenate all samples
df.0 <- data.frame()

for (i in 1:nrow(df.meta)) {
  df.0 <- rbind(df.0, load_variant_table(path, df.meta$project[i], df.meta$sample[i]))
}

# append metadata
df.0 <- merge(df.0, df.meta, all.x=T, all.y=F)

# split mutation signature
df.0$mut.split <- sapply(df.0$mutation, function(x) strsplit(as.character(x),"[S:>]"))
df.0$mutation.pos <- as.integer(sapply(df.0$mut.split, function(x) x[2]))
df.0$mutation.from <- sapply(df.0$mut.split, function(x) x[3])
df.0$mutation.to <- sapply(df.0$mut.split, function(x) x[4])
df.0$mut.split  <- NULL

# mutation signature groups
df.0$mutation.signature <- paste(df.0$mutation.from, df.0$mutation.to, sep =">")
df.0 <- merge(df.0, sign.rep, all.x=T, all.y=F)

# exclude bases that are not in reference
df.0 <- subset(df.0, !(mutation.pos %in% 22:25))

# shift back mutations
df.0$mutation.pos <- ifelse(df.0$mutation.pos > 25, df.0$mutation.pos - 4, df.0$mutation.pos)

df.0$mutation <- with(df.0, paste("S", mutation.pos, ":", mutation.from, ">", mutation.to, sep=""))
  
df <- subset(df.0, project %in% c("polerr2016-1", "polerr2016-2"))
df.linpcr <- subset(df.0, project %in% c("polerr73", "polerr82"))

template <- "TAGCGTGAAGACGACAGAACCATGGGATCCATTATCGGCGGCGAATTTACCACCATTGAAAACCAGCCGTGGTTTGCGGCGATTTATCGTCGTCATCGTGGCGGCAGCGTGACCTATGTGTGCGGCGGCAGCCTGATTAGCCCGTGCTGG" # 4 index bases removed
```

## Estimating the error rates

In this section we compute linear PCR error rate from Proj73/82 and error rate in conventional PCR from Polerr2016 project. Note that the linear PCR error rate is substantially higher than rate per cycle of conentional PCR. For some cases, e.g. Phusion it accounts for ~1/2 of errors observed in Polerr2016.

```{r}
# Compute linear PCR error rate

df.er.linpcr <- ddply(df.linpcr, .(project, name), summarize, mismatches = sum(count.major), umi.count = mean(coverage))
df.er.linpcr <- ddply(df.er.linpcr, .(name), summarize, mismatches = sum(mismatches), umi.count = round(sum(umi.count)))

df.er.linpcr <- data.frame(name = df.er.linpcr$name, linpcr.er = df.er.linpcr$mismatches/df.er.linpcr$umi.count/nchar(template))

# Compute uncrorrected error rate

df.er <- ddply(df, .(project, name, cycles), summarize, 
               mismatches = sum(count.major), umi.count = round(mean(coverage)))

df.er <- merge(df.er, df.er.linpcr, by = "name", all.x = T)

df.er$err.rate <- with(df.er, mismatches / umi.count / nchar(template) / mean(cycles))
df.er$delta <- with(df.er, sqrt(mismatches / umi.count * (1 - mismatches / umi.count) / umi.count) / 
                 nchar(template) / mean(cycles))
df.er$err.lb <- df.er$err.rate - 1.96 * df.er$delta
df.er$err.ub <- df.er$err.rate + 1.96 * df.er$delta

# Error rates corrected for linear PCR errors

df.er$mismatches.corr <- with(df.er, mismatches - linpcr.er * umi.count * nchar(template))
df.er$err.rate.corr <- with(df.er, mismatches.corr / umi.count / nchar(template) / mean(cycles))
df.er$delta.corr <- with(df.er, sqrt(mismatches.corr / umi.count * (1 - mismatches.corr / umi.count) / umi.count) / 
                 nchar(template) / mean(cycles))
df.er$err.lb.corr <- df.er$err.rate.corr - 1.96 * df.er$delta.corr
df.er$err.ub.corr <- df.er$err.rate.corr + 1.96 * df.er$delta.corr

df.er$cycles <- NULL

kable(df.er)

write.table(df.er, file="er.txt", quote=F, sep="\t", row.names = F)
```

## Patterns and recurrent errors

Substitution signature preferences:

```{r}
df.pattern <- ddply(df, .(name, mutation.signature.rep), summarize, count.sum = sum(count.major))
df.pattern <- ddply(df.pattern, .(name), transform, freq = count.sum / sum(count.sum))

ggplot(df.pattern, aes(x = mutation.signature.rep, weight = freq, 
               fill = name)) + geom_bar(position = position_dodge()) +
  xlab("Substitution signature") + scale_y_continuous("Share of errors", expand=c(0,0)) +
  scale_fill_brewer(palette = "Paired") + theme_bw()

# same for linpcr

df.pattern.linpcr <- ddply(df.linpcr, .(name, mutation.signature), summarize, count.sum = sum(count.major))
df.pattern.linpcr <- ddply(df.pattern.linpcr, .(name), transform, freq = count.sum / sum(count.sum))

ggplot(df.pattern.linpcr, aes(x = mutation.signature, weight = freq, 
               fill = name)) + geom_bar(position = position_dodge()) +
  xlab("Substitution signature") + scale_y_continuous("Share of errors", expand=c(0,0)) +
  scale_fill_brewer(palette = "Paired") + theme_bw()
```

Clustering of polymerase error profiles:

```{r}
df$freq <- df$count.major / df$coverage
df$name_proj <- paste(df$name, ifelse(df$project == "polerr2016-1", "1", "2"), sep="-")
mat.profile <- dcast(df, name_proj ~ mutation, value.var = "freq")
rownames(mat.profile) <- mat.profile[,1]
mat.profile[,1] <- NULL
mat.profile <- t(as.matrix(mat.profile))
mat.profile[is.na(mat.profile)] <- 0

df.aux <- unique(data.frame(mutation = df$mutation, signature = df$mutation.signature.rep))
df.aux$mutation <- as.character(df.aux$mutation)

df.color <- merge(data.frame(mutation = rownames(mat.profile)), df.aux)

df.color <- merge(df.color, data.frame(colors = brewer.pal(6, "Paired"), signature = levels(df.color$signature)))

rowcolor <- as.character(df.color$colors)
names(rowcolor) <- df.color$mutation

rowDend <- hclust(dist(mat.profile))
colDend <- hclust(as.dist((1-cor(mat.profile, method = "spearman"))/2), method = "ward")

heatmap.2(log10(mat.profile), col=c("#67001f", colorpanel(99, "#b2182b", "#f7f7f7", "#2166ac")),
          dendrogram = "both", RowSideColors = rowcolor, breaks = seq(-6,-2,length.out = 101),
          Rowv = as.dendrogram(rowDend), 
          Colv = as.dendrogram(colDend),
          density.info = "none", trace="none")
```

Recurrent errors

```{r}
df.1x <- ddply(subset(df, project == "polerr2016-1"), .(name, mutation), summarize,
               freq = count.major / coverage, coverage = coverage)

df.1y <- ddply(subset(df, project == "polerr2016-2"), .(name, mutation), summarize,
               freq = count.major / coverage, coverage = coverage)

df.1 <- merge(df.1x, df.1y, by=c("name","mutation"), all = T)
mask1 <- is.na(df.1$freq.x)
mask2 <- is.na(df.1$freq.y)
df.1$miss <- mask1 | mask2

df.1 <- ddply(df.1, .(name), transform, 
              freq.x = ifelse(is.na(freq.x), 1/mean(coverage.x, na.rm = T), freq.x),
              freq.y = ifelse(is.na(freq.y), 1/mean(coverage.y, na.rm = T), freq.y))

ggplot() + 
  geom_point(data=subset(df.1, !miss), aes(x=freq.x, y=freq.y), shape=21) + 
  geom_smooth(data=subset(df.1, !miss & name != "Phusion"), aes(x=freq.x, y=freq.y), method="lm") + 
  geom_point(data=subset(df.1, miss), aes(x=freq.x, y=freq.y), shape=21, color="red") + 
  facet_wrap(~name) +
  scale_x_log10("Frequency in experiment 1", limits=c(1e-6,1e-2)) + 
  scale_y_log10("Frequency in experiment 2", limits=c(1e-6,1e-2)) + 
  theme_bw()
```

Frequency distribution of individual mutations, grouped by their pattern.

```{r}
df.mean.err <- ddply(df, .(name), summarize, mean.err.rate = sum(count.major) / mean(coverage) / nchar(template))
# ^ this one is the gloal mean

df <- merge(df, df.mean.err, all.x=T, all.y=F)

ggplot(df) + 
  geom_histogram(aes(x = count.major / coverage, fill = mutation.signature.rep)) + 
  geom_linerange(aes(x = mean.err.rate, ymin = 0, ymax=50), linetype = "dashed", color="black") +
  scale_fill_brewer("Substitution signature", palette = "Paired") +
  scale_x_log10("Error frequency") + 
  scale_y_continuous("", expand=c(0,0)) + facet_wrap(~name) + theme_bw()
  #facet_wrap(~name, scales = "free_y") + theme_bw()
```

## Sequence quality and error patterns

```{r}
df.q <- read.table("data/mmqc.txt", header=T,sep="\t")
df.q <- melt(df.q, id.vars = "qual.threshold")
df.q$variable <- as.factor(gsub(".", ">", df.q$variable, fixed = T))
colnames(df.q) <- c("qual.threshold", "mutation.signature", "freq")

# Error profiles for different quality threhsolds
ggplot(df.q, aes(x=qual.threshold, color=mutation.signature, y=freq)) + 
  geom_smooth() + geom_point() + scale_y_log10() +
  scale_color_brewer(palette = "Paired") + theme_bw()

# Correlation between error profiles

df.linpcrT <- ddply(df.linpcr, .(mutation.signature, name), summarize, freq = sum(count.major) / sum(coverage))
df.q_truseq <- merge(df.q, df.linpcrT, by = c("mutation.signature"), all.x=T)
df.q_truseq$freq.y[is.na(df.q_truseq$freq.y)] <- 0

df.q_truseq <- ddply(df.q_truseq, .(qual.threshold, name), summarize, R = cor(freq.x, freq.y))

ggplot(df.q_truseq, aes(x=qual.threshold, y=R, color=name)) + 
  geom_line() + geom_point(shape=21) + scale_color_brewer("Sample", palette = "Paired") +
  scale_y_continuous("Pearson correlation", limits=c(-1,1)) + xlab("Quality threshold, Phred") +
  theme_bw()
```

## Supplementary data

Error rate consistency between two independent experimental replicas from Polerr2016 dataset.

```{r}
df.er.cast <- dcast(df.er, name ~ project, 
                    value.var = "err.rate.corr")
df.er.cast2 <- dcast(df.er, name ~ project, 
                    value.var = "delta")

df.er.cast <- merge(df.er.cast, df.er.cast2, by = "name")
colnames(df.er.cast) <- c("name", "replica1.x", "replica2.x", "replica1.y", "replica2.y")

m <- lm(replica1.x ~ replica2.x, df.er.cast);
eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(R)~"="~r,
     list(a = format(coef(m)[1], digits = 2),
          b = format(coef(m)[2], digits = 2),
          r = format(sqrt(summary(m)$r.squared), digits = 2)))
lbl<-as.character(as.expression(eq))

ggplot(df.er.cast, aes(x=replica1.x, y=replica2.x)) + 
  geom_errorbarh(aes(xmax=replica1.x+1.96*replica1.y,xmin=replica1.x-1.96*replica1.y)) +
  geom_errorbar(aes(ymax=replica2.x+1.96*replica2.y,ymin=replica2.x-1.96*replica2.y)) +
  geom_smooth(method = "lm", color="black", fill="grey80", fullrange = T) + 
  geom_point(size=2, shape=21, fill="grey70") + 
  geom_text(aes(label=name), vjust=1, hjust = .3, color="red") + 
  geom_label(aes(x = 1e-6, y = 8e-5, label = lbl), hjust=-0.1, parse = TRUE)+
  scale_x_log10(name="Error rate in experiment 1, 1/bp/cycle", limits=c(1e-6,1e-4)) + 
  scale_y_log10(name="Error rate in experiment 2, 1/bp/cycle", limits=c(1e-6,1e-4)) +  
  theme_bw()
```

Error distribution by position

```{r}
df.pos <- ddply(df, .(name, mutation.pos, mutation.from, mutation.signature.rep), 
                summarize, count = sum(count.major), freq = sum(count.major) / coverage[1])

ggplot(df.pos, aes(x = mutation.pos, y = freq, color = mutation.signature.rep)) + 
  geom_line() + geom_point(shape=21, fill="white") + 
  scale_color_brewer("Substituted base", palette = "Paired") +
  xlab("Position on template") + scale_y_log10("Error rate") +
  facet_grid(name~mutation.signature.rep, scales="free_y") +
  theme_bw()
```

UMI coverage histogram

```{r}
df.meta <- subset(df.meta, project %in% c("polerr2016-1", "polerr2016-2"))

df.h <- data.frame(mig.size.bin = integer(), read.count = integer(), name=character(), project=character())

for (proj in unique(df.meta$project)) {
  for (sample in unique(subset(df.meta, project == proj)$sample)) {
    mask <- which(df.meta$sample == sample & df.meta$project == proj)
    name <- df.meta$name[mask][1]
    cycles_2 <- df.meta$cycles_2[mask][1]
    df.hh <- read.table(paste(path, paste(proj, sample, "umi.histogram.txt", sep="."), sep="/"), 
                        header=T, sep="\t")
    df.h <- rbind(df.h, data.frame(mig.size.bin = df.hh$mig.size.bin, read.count = df.hh$read.count, 
                                   name=name, project=proj, cycles_2=cycles_2))
  }
}

ggplot(df.h) + 
  geom_rect(aes(xmin=1, xmax=5, ymin=0, ymax=Inf), fill="grey") +
  geom_line(aes(x=mig.size.bin, y=read.count, color=project, linetype=name, 
                group = interaction(project, name)), size=1) + 
  scale_x_log10("UMI coverage", limits = c(1,1e6), expand=c(0,0)) + 
  scale_y_continuous("Number of reads in UMIs", expand=c(0,0), limits=c(0,1.5e7)) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()
```

Coverage and PCR cycles

```{r}
df.o <- ddply(df.h, .(project, name, cycles_2), summarize, 
              peak = log2(mig.size.bin[which(read.count == max(read.count))]))

m <- lm(peak ~ cycles_2, df.o)
eq <- substitute(italic(R)~"="~r*","~~italic(P)~"="~p,
     list(
     r = format(sqrt(summary(m)$r.squared), digits = 2),
     p = format(summary(m)$coefficient[[8]], digits = 3)
     ))
lbl<-as.character(as.expression(eq))

ggplot()+
geom_label(aes(x = 22, y = 12, label = lbl), hjust=-0.1, parse = TRUE)+
stat_smooth(data=df.o, aes(cycles_2, peak), method=lm, color="black", fill="grey80") + 
geom_jitter(data=df.o, aes(cycles_2, peak, fill=name), size=3, width=0.3, height=0.3, color="black", shape=21) +
scale_x_continuous(name="2nd PCR cycles", breaks=10:30) +
scale_y_continuous(expression('log'[2]~'characteristic MIG size'), breaks=2:20) +
scale_fill_brewer(name  ="Polymerase", palette = "Paired") +
theme_bw()

```